---
title: "Homework 1"
author: "Jack Cunningham"
format: pdf
editor: visual
---

2.1)

```{r,echo=FALSE}
occupation_data = matrix(c(.018,.002,.001,.001,.001,.035,
                           .112,.066,.018,.029,.031,.064,.094,.019,
                           .032, .008, .032, .032,.010, .043, .018,
                           .069,.084, .051, .13),nrow = 5, ncol = 5)

occupation_labels <- c("farm","operatives","craftsmen","sales",
                               "professional")

rownames(occupation_data) <- c("farm","operatives","craftsmen","sales",
                               "professional")
colnames(occupation_data) <- c("farm","operatives","craftsmen","sales",
                               "professional")

```

a\)

$$
p_{Y_{1}}(y_1)= \sum_{y_2 \in Y_2} p_{Y_1,Y_2}(y_1,y_2)
$$

Which leaves us with the marginal probability distribution of the father being:

```{r}
father_marginal_distribution <- rowSums(occupation_data)
names(father_marginal_distribution) <- occupation_labels
father_marginal_distribution
```

b\)

Similarly we have:

$$
p_{Y_2}(y_2)=\sum_{y_{1} \in Y_1}p_{Y_1,Y_2}(y_1,y_2)
$$

Which leaves us with the marginal probability distribution of the son being:

```{r}
son_marginal_distribution <- colSums(occupation_data)
names(son_marginal_distribution) <- occupation_labels
son_marginal_distribution
```

c\)

The conditional distribution of a son's occupation given that the father is a farmer is:

$$
p_{Y_2|Y_{1}}(y_2|y_1=\text{farmer})=\frac{Pr(\{Y_{1}=\text{farmer\} } \cap\{Y_{2}=y_2\})}{Pr(Y_{1} = \text{farmer})}
$$

```{r}
occupation_data["farm",]/father_marginal_distribution["farm"]
```

d\)

The conditional distribution of a father's occupation given that the son is a farmer is:

$$
p_{Y_1|Y_2}(y_1|y_{2}= \text{farmer})=\frac{Pr(\{Y_{1}=y_1 \}\cap \{ Y_{2}=\text{farmer}\}}{Pr(Y_2 = \text{farmer})}
$$

```{r}
occupation_data[,"farm"]/son_marginal_distribution["farm"]
```

2.2)

Since the two random variables are independent we have:

a\)

$$
E[a_1Y_1+a_2Y_2]=a_1 \mu_1+a_2\mu_2 \quad Var(a_1 Y_1+a_2 Y_2)=a_1^2 \sigma_1^2+a_2^2+\sigma_2^2
$$

b\)

$$
E[a_1Y_1-a_2Y_2]=a_1 \mu_1-a_2\mu_2 \quad Var(a_1 Y_1-a_2 Y_2)=a_1^2 \sigma_1^2+a_2^2+\sigma_2^2
$$
